---
title: "01_metadata_setup"
author: "Tristan Caro"
format: html
editor: visual
---

This code only needs to be run once to create a template metadata file that will be manually annotated and QC'd. The code should be re-run if any changes to the raw `.avg` or `.mu` files are made, so that the `.rds` cache file is updated.

## Setup

```{r}
library(tidyverse)
source("source/read_mu_file.R")
```

## Import the data

Create a function for importing:

```{r}
import_many_mu_files <- function(directory) {
  directory |> 
  map(read_mu) |> 
  bind_rows(.id = "file")
} 

```

Import Dec 2022 data

```{r}
dec22_fp <- "data/bl143_dec22/02_mu/"
dec22_files <- list.files(path = dec22_fp, pattern = "\\.mu$", full.names = TRUE)
dec22_data <- import_many_mu_files(dec22_files) |> 
  mutate(run = "dec_2022")
```

Import Dec 2023 data

```{r}
dec23_fp <- "data/bl143_dec23/02_mu/"
dec23_files <- list.files(path = dec23_fp, pattern = "\\.mu$", full.names = TRUE)
dec23_data <- import_many_mu_files(dec23_files)|> 
  mutate(run = "dec_2023")
```

Import Jun 2024 data

```{r}
jun24_fp <- "data/bl143_jun24/02_mu/"
jun24_files <- list.files(path = jun24_fp, pattern = "\\.mu$", full.names = TRUE)
jun24_data <- import_many_mu_files(jun24_files) |> 
  mutate(run = "jun_2024")
```

Combine into one big tibble

```{r}
full_dataset <- bind_rows(
  dec22_data, dec23_data, jun24_data
)

saveRDS(full_dataset, file = "cache/xanes_data_raw.RDS")
```

Export a list of all files to build metadata

```{r}
full_metadata <- full_dataset |> 
  distinct(filename) |> 
  mutate(
    core = case_when(
      str_detect(filename, "BA1B") ~ "BA1B",
      str_detect(filename, "BA4A") ~ "BA4A",
      str_detect(filename, "BA3A") ~ "BA3A",
      TRUE ~ NA
    ),
    is_TLE = case_when(
      str_detect(filename, "TLE") ~ TRUE,
      TRUE ~ FALSE
    )
  )
```

Write the full metadataset to an excel file

```{r}
full_metadata |> writexl::write_xlsx(path = "data/metadata.xlsx")
```
